---
title: "R Notebook"
output: html_notebook
---

Determine the parties: Identify the political parties that contested in each election. This information is available in the "party" column of the dataset.

```{r}
library(dplyr)
data = read.csv("/Users/rahulgupta/Dropbox (Dartmouth College)/Mac/Documents/Dartmouth/FYREE/Political Polarization/countypres.csv")
df <- read.csv("/Users/rahulgupta/Dropbox (Dartmouth College)/Mac/Documents/Dartmouth/FYREE/Political Polarization/countypres.csv")

# Select only the necessary columns
df <- select(df, year, state, county_name, party, candidatevotes, totalvotes)

# Get the unique parties for each election in each county
parties <- df %>%
  group_by(year, state, county_name) %>%
  distinct(party) %>%
  filter(!is.na(party)) %>%
  pull(party)

# Print the unique parties
unique(parties)
```


Calculate the vote share: Calculate the vote share of each political party for each election in each county. You can do this by dividing the number of votes received by each party by the total number of votes cast in the county for that election.

```{r}
# Calculate the vote share for each party in each county for each election
df <- df %>%
  group_by(year, state, county_name, party) %>%
  summarise(total_votes = sum(candidatevotes)) %>%
  ungroup() %>%
  group_by(year, state, county_name) %>%
  mutate(total_county_votes = sum(total_votes),
         vote_share = total_votes/total_county_votes) %>%
  select(-total_votes, -total_county_votes)

# Print the first few rows of the resulting dataset
head(df)
```


Calculate the polarization score: Once you have the vote share data, you can calculate a polarization score for each county for each election. There are several ways to do this, but one possible method is to use the Gini coefficient. The Gini coefficient is a measure of inequality, but it can also be used to measure polarization. A higher Gini coefficient indicates a more polarized county. You could also explore other measures such as the Euclidean distance or Kullback-Leibler divergence.

```{r}
# Calculate the polarization score using the Gini coefficient
library(ineq)

df_polarization <- df %>%
  group_by(year, state, county_name) %>%
  summarise(gini_coef = 1 - ineq(vote_share, type = "Gini"))

# Print the first few rows of the resulting dataset
head(df_polarization)

```


Analyze the results: Once you have calculated the polarization score for each county for each election, you can analyze the results to identify trends and patterns. You could use visualizations such as heat maps or scatter plots to visualize the data and identify any patterns.

Heat Map
```{r}
# Filter the data for the 2020 election
df_polarization_2020 <- df_polarization %>%
  filter(year == 2020)

# Create a heat map of the polarization scores
library(ggplot2)
ggplot(df_polarization_2020, aes(x = county_name, y = state, fill = gini_coef)) +
  geom_tile() +
  scale_fill_gradient(low = "red", high = "white") + # reverse the order of the color gradient
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("Polarization Scores for US Counties in the 2020 Presidential Election")


```


Scatter Plot
```{r}
# Filter the data for the 2016 and 2020 elections
df_polarization_16_20 <- df_polarization %>%
  filter(year %in% c(2016, 2020))

# Create a scatter plot of the polarization scores
library(ggplot2)

ggplot(df_polarization_16_20, aes(x = gini_coef, y = year, color = state)) +
  geom_point(size = 3) +
  scale_color_manual(values = scales::hue_pal()(51)) +
  facet_wrap(~state, ncol = 5, scales = "free_x") +
  theme(legend.position = "none") +
  labs(x = "Polarization Score (Gini Coefficient)", y = "Year",
       title = "Polarization Scores for US Counties in the 2016 and 2020 Presidential Elections")

```

Plot on the US Map
```{r}
df_polarization_2020 <- merge(df_polarization_2020, data)
results <- rename(df_polarization_2020, fips = county_fips)

# Generate color palette
library(RColorBrewer)
colors <- brewer.pal(n = 6, name = "Paired")
library(usmap)
plot_usmap(data = results,
           values = "gini_coef",
           color = "white") +
  scale_fill_gradientn(colors = colors,
                       name = "Political Polarization",
                       label = scales::comma) +
  labs(title = "United States",
       subtitle = "Presidential Election Data from 2000 to 2020") +
  theme(legend.position = "right")
```

